Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-10-17 Manuscript/SS_GAM.tex
Upload tex file

\address[a]{Department of Biostatistics, University of Alabama at
Birmingham, Birmingham, USA}
\address[b]{Department of Biostatistics and Data Science, Wake Forest
University, Winston-Salem, USA}

@@

\abstract{There are proposals that extend the classical generalized
additive models (GAMs) to accommodate the high-dimensional (\(p>>n\))
problem using group penalties and priors. These models rarely smooth the
nonlinear effect properly as the sparse regularization leads to excess
shrinkage on coefficients and inaccurate prediction. Moreover, most of
these GAMs consider an ``all-in-all-out'' approach for function
selection, rendering them difficult to answer if nonlinear effects are
necessary. While some Bayesian models can address these shortcomings,
using Markov chain Monte Carlo (MCMC) algorithms for model fitting
creates a new challenge, scalability. Hence, we propose Bayesian
hierarchical generalized additive models as a solution: we propose a
novel spike-and-slab spline prior for smoothing functions, where a
reparameterization step facilitates smoothing penalty and separate
selection of linear and nonlinear spaces. Two scalable and deterministic
algorithms, EM-Coordinate Descent and EM-Iterative Weighted Least
Squares, are developed to fit the proposed model. Simulation studies and
metabolomics data analyses demonstrate improved performance against
state-of-the-art models, mgcv, COSSO and sparse Bayesian GAM. The
software implementation of the proposed models is freely available via
an R package BHAM.}

@@

Many modern biomedical research, e.g.~sequencing data analysis, electric
health record data analysis, require special treatment of
high-dimensionality, commonly known as \(p >> n\) problem. There is
extensive literature on high-dimensional linear models via penalized
models or Bayesian hierarchical models, see Mallick and Yi
\citep{Mallick2013} for review. These models are built upon a
restrictive and unrealistic assumption, linearity. In classical
statistical modeling, many strategies and models are proposed to relax
the linearity assumption with various degrees of complexity. For
example, variable categorization is a simple and common practice in
epidemiology, but suffers from power and interpretation issues. More
complex models to address nonlinear effects include random forest and
other so-called ``black box'' models \citep{Breiman2001}. These models
are not particularly attractive in biomedical studies, as they lack the
transparency to understand the etiology of diseases. In addition, how to
generalize these models to the high-dimensional setting remains unclear.

Nonparametric regression models are appropriate alternatives to the
``black-box'' models thanks to their balance between model flexibility
and interpretability. Among those, generalized additive models (GAMs),
proposed in the seminal work of Hastie and Tibshirani
\citep{Hastie1987}, grew to be one of the most popular modeling tools.
In a GAM, the response variable, \(Y\), which is assumed to follow some
exponential family distribution, denoted by \(EF(\mu, \phi)\), with mean
\(\mu\) and dispersion \(\phi\), can be modeled with the summation of
smoothing functions, \(B_j(\cdot), j = 1, \dots, p\), of a given
\(p\)-dimensional vector of covariates \(\boldsymbol{x}\), written as \[
 E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)),
\] where \(g^{-1}(\cdot)\) is the inverse of a monotonic link function.
The smoothing functions can take many forms and are estimated using a
pseudo-weighted version of the backfitting algorithm
\citep{Breiman1985}. Nevertheless, the classical GAMs cannot fulfill the
increasing analytic demands for high-dimensional data analysis in
biomedical studies.

There exists some proposals to generalize the classical GAM to
accommodate high-dimensional applications. The regularized models,
branching out from group regularized linear models, are used to fit GAMs
by accounting for the structure introduced when expanding smoothing
functions. Ravikumar et al. \citep{Ravikumar2009} extended the grouped
lasso \citep{Yuan2006} to additive models (AMs); Huang et al.
\citep{Huang2010} further developed adaptive grouped lasso for additive
models; Wang et al. \citep{Wang2007} and Xue \citep{Xue2009}
respectively applied grouped SCAD penalty \citep{Fan2001} to additive
models. Recent Bayesian hierarchical models are also used in the context
of high-dimensional additive models. Various group spike-and-slab priors
combining with computationally intensive Markov chain Monte Carlo (MCMC)
algorithms \citep{Xu2015, Yang2020} are proposed, where the application
on AMs are by-products. Bai et al. \citep{Bai2020} was the first to
apply group spike-and-slab lasso prior to Gaussian AMs using a fast
optimization algorithm, and further generalized the framework to
GAMs\citep{Bai2021}. As these methods focus on addressing the sparsity,
it is very likely to overly penalize the basis function coefficients and
produce inaccurate predictions and curve interpolation, particularly
when complex signals are assumed and large number of knots are used.
\citep{Scheipl2013} In addition, these methods adapt an `all-in-all-out'
strategy, i.e.~either including or excluding the variable completely,
rendering no space for bi-level selection. Scheipl et al.
\citep{Scheipl2012} proposed a spike-and-slab structure prior,
normal-mixture-of-inverse gamma prior, that address the previous
challenges. But the model fitting relies on computational intensive MCMC
algorithmic and creates scalability concern. It would be of special
interest to develop a fast, flexible and accurate generalized additive
model framework.

@@

The proposed framework, BHAM, differs from previous spike-and-slab based
GAMs, i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB-GAM
\citep{Bai2021} in many ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM.
Spike-and-slab lasso priors provide computational convenience during
model fitting by using optimization algorithms instead of computational
intensive sampling algorithms. They make fitting high- and
ultrahigh-dimensional models more feasible without sacrificing
performance in prediction and variable selection. Secondly, the
optimization algorithms introduced in this article include EM-Coordinate
Descent (EM-CD) algorithm and EM-Iterative weighted least square
(EM-IWLS) algorithm for different utilities. The applications of the two
aforementioned algorithms are fruitful in Bayesian hierarchical models
for high-dimensional data. \citep{Yi2012, Rockova2014a, Rockova2018}
Even though SB-GAM with a group spike-and-slab lasso prior also uses
EM-CD for model fitting, the proposed EM-IWLS provides uncertainty
measures that could be further used for confidence bound calculation and
hypothesis testing. EM-IWLS complements the lack of inference potential
from EM-CD in a median and high dimensional setting. Furthermore, the
proposed model addresses the incapability of bi-level selection in
SB-GAM. Lastly, the R implementation of the Bayesian hierarchical
additive model is available publicly via BHAM at
\url{https://github.com/boyiguo1/BHAM}, making transnational science
more accessible.

In Section \ref{sec:BHAM}, we establish the Bayesian hierarchical
generalized additive model, introduce the proposed spike-and-slab spline
priors, and describe the two fast-fitting algorithms. In Section
\ref{sec:sim}, we compare the proposed framework to state-of-art models,
mgcv, COSSO and sparse Bayesian GAM via Monte Carlo simulation studies.
Analyses of two metabolomics datasets are presented in Section
\ref{sec:real_data}. Conclusion and discussions are given in Section
\ref{sec:concl}.

@@

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} and the data distribution is \begin{equation}
f(\boldsymbol{Y} = \boldsymbol{y}| \boldsymbol{\beta}, \phi) = \prod\limits^n_{i=1}f( Y = y_i|\boldsymbol{\beta}, \phi),\nonumber
\end{equation} with smoothing functions \(B_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(B_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+3\)
elements column vector. Its corresponding smoothing penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable such that
\(B_j(x_j) = \boldsymbol{\beta}_j^T \boldsymbol{X}_j\). With slight
abuse of notation, we denote vectors and matrices in bold fonts
\(\boldsymbol{\beta}, \boldsymbol{X}\) with conformable dimensions,
where scalar and random variables are denoted in with unbold fonts
\(\beta, X\). The matrix transposing operation is denoted with a
superscript \(^T\). To note, the proposed model can include parametric
forms of variables in the model, treating the parametric function as a
special case of the smoothing function, e.g.~\(B_j(x_j) = x_j\) with the
smoothing penalty matrix defined as
\(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

@@

The re-parameterization step sets up the foundation of the proposed
model and provides three-fold benefits. First of all, the
re-parameterization integrates the smoothing penalty into the design
matrix, and encourages models to properly smooth the nonlinear function
in addition to the sparse penalty for function selection. Secondly, the
eigendecomposition of the smoothing penalty allows the isolation of the
linear from the nonlinear space, improving the feasibility of bi-level
function selection and inference. The eigendecomposition facilitates the
construction of orthonormal design matrix, which makes imposing
independent priors on the coefficients possible. This reduces the
computational complexity compared to using a multivariate priors, and
greatly broadens the choices of priors and further model choices. Last
but not least, the scaling step of \(\boldsymbol{X}^*\) further
simplifies the choice of priors, from column-specific priors to a
unified prior.

\hypertarget{spike-and-slab-smooth-priors}{%
\subsection{Spike-and-Slab Smooth
Priors}\label{spike-and-slab-smooth-priors}}

The family of spike-and-slab (SS) priors regression models dominates
Bayesian high-dimensional analysis for its utility in outcome prediction
and variable selection. We defer to Bai et al. \citep{Bai2021Review} for
an in-depth introduction to spike-and-slab priors. To summarize,
spike-and-slab priors are a family of mixture distributions that
comprises a skinny spike density \(f_{\text{spike}}(\cdot)\) for weak
signals and a flat slab density \(f_{\text{slab}}(\cdot)\) for strong
signals, mathematically \[
 \beta|\gamma \sim (1-\gamma)f_{\text{spike}}(\beta) + \gamma f_{\text{slab}}(\beta).
\] The most distinct feature of SS priors is that it is conditioned on a
latent binary variable \(\gamma \in \{0,1\}\) that indicates whether the
variable \(x\) is included in the model. There are various
spike-and-slab priors depending on the choice for the spike density
\(f_{\text{spike}}(\cdot)\) and the slab density
\(f_{\text{slab}}(\cdot)\), see George and McCulloch
\citep{George1993, George1997}; Chipman \citep{Chipman1996} for grouped
variables; Brown et al. \citep{Brown1998} for multivariate outcomes;
Ishwaran and Rao \citep{Ishwaran2005}; Clyde and George
\citep{Clyde2004} and reference therein.

The major criticism of early spike-and-slab models is being
computationally prohibitive. \citep{Bai2021Review} Since then, many
studies focus on alleviating the computational burden that sampling
algorithms bear, which include EMVS \citep{Rockova2014a} and
spike-and-slab Lasso \citep{Rockova2018b, Rockova2018}. Particularly,
the development of spike-and-slab Lasso substantially improves the
scalability of SS models, setting up the theoretical foundation for
generalized models in -omics data analysis
\citep{Tang2017a, Tang2017, Tang2018, Tang2019}. The spike-and-slab
Lasso (SSL) prior is composed of two double exponential distributions
with mean 0 and different dispersion parameters, \(0 < s_0 < s_1\),
mathematically, \begin{equation} 
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber
\end{equation} Given that both double exponential distributions have a
mean of 0 and the latent indicator \(\gamma\) can only take the value of
0 or 1, the mixture double exponential distribution can be formulated as
one single double exponential density, \begin{equation} \label{eq:ssl}
\beta | \gamma \sim DE(0, S), 0 < s_0 < s_1,
\end{equation} with the scale parameter
\(S = (1-\gamma)s_0 + \gamma s_1\). The SSL also mitigates the problem
of EMVS where the weak signals are not shrink to zero, and hence is
preferred in high-dimensional data analysis. We notice that Bai
\citep{Bai2021} is the first to apply spike-and-slab lasso prior in the
GAM framework, where the densities of the spike and slab components take
the group lasso density \citep{Xu2015}. Bai \citep{Bai2021} takes an
``all-in-all-out'' strategy for function selection.

\hypertarget{spike-and-slab-lasso-prior}{%
\subsubsection{Spike-and-Slab Lasso
Prior}\label{spike-and-slab-lasso-prior}}

We introduce a novel prior for GAMs, particularly for high-dimensional
nonlinear modeling with bi-level selection. The proposed prior extends
from the spike-and-slab lasso prior described in Equation
(\ref{eq:ssl}). Given the re-parameterized deign matrix
\(\boldsymbol{X}_j = \begin{bmatrix} X^0_j : \boldsymbol{X}_j^*\end{bmatrix}\)
for the \(j\)th variable, we impose a two-part SSL prior to the
coefficients
\(\boldsymbol{\beta}_j = \begin{bmatrix} \beta_j : \boldsymbol{\beta}_j^*\end{bmatrix}\).
Specifically, we impose independent group priors on the linear space
coefficients and on the nonlinear space coefficients respectively,
\begin{align}\label{eq:bham_ssl}
  \beta_{j} |\gamma_{j},s_0,s_1 &\sim DE(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1) \nonumber \\
  \beta^*_{jk} | \gamma^*_{j},s_0,s_1 &\overset{\text{iid}}{\sim}DE(0,(1-\gamma^*_{j}) s_0 + \gamma^*_{j} s_1), k=1,\dots, K_j
\end{align} where \(\gamma_{j}\in\{0,1\}\) and
\(\gamma^*_{j}\in \{0,1\}\) are two latent indicator variables,
indicating if the model includes the linear effect and the nonlinear
effect of the \(j\)th variable respectively. \(s_0\) and \(s_1\) are
scale parameters, assuming \(0 < s_0 < s_1\) and given. These scale
parameters \(s_0\) and \(s_1\) can be treated as tuning parameters and
optimized via cross-validation. A discussion of how to choose the scale
parameters comes in Section \ref{sec:tune}. To note, this prior differs
from previous group spike-and-slab lasso priors
\citep{Tang2018, Tang2019}, as the \(\beta_j\) and
\(\boldsymbol{\beta}^*_j\) have different indicator variables
\(\gamma_j\), \(\gamma_j^*\) respectively. It is possible to add a more
restrictive assumption on the priors, assuming that one indicator
variable decides the inclusion of both the linear effect and nonlinear
effect, i.e.~\(\gamma_j = \gamma^*_j\). This converges to the SB-GAM
\citep{Bai2021}. Conversely, it is also possible to relax the assumption
such that each coefficient \(\beta_{jk} \in \boldsymbol{\beta}^*_j\) has
its own latent indicator \(\gamma_{jk}\), but at the cost of
complicating the bi-level function selection. This reduces the proposed
prior to the classic spike-and-slab lasso prior.

The re-parameterization introduced in Section \ref{sec:BHAM} grants the
validity of the proposed prior. First of all, the smoothing function
bases are linear dependent and ask for extra attention. The
eigeondecomposition remedies the problem and hence our prior can be set
to be conditionally independent. Secondly, the eigenvalue scaling
provides a panacea to allow unified scale parameters for all bases of
all smoothing functions of variables.

@@

\caption{Directed acyclic grpah of BHAM-SSL model with parameter expansion. Elliposes are stochastic nodes, rectangles and are deterministic nodes. }
\label{fig:SSprior}
\end{figure}

@@

In this section, we extend the two EM-based algorithms, EM-CD and
EM-IWLS algorithms, to fit BHAMs. To note, the two proposed algorithms
provides different utilities. The EM-CD algorithm is specifically for
fitting BHAM-SSL with an expedited performance, recommending to use in
high and ultra-high dimensional setting. A specific concern of EM-CD
algorithm is that it provides no information for inference. In contrast,
the EM-IWLS can extract the variance-covariance matrix and be used to
construct Wald type inference. Moreover, the EM-IWLS is a more general
model fitting algorithm that can be used for fitting not only SSL and
continuous SS priors but also Student's t-priors and double exponential
priors. To note, SB-GAM\citep{Bai2020, Bai2021} also used an EM-CD
algorithm. The main difference between proposed EM-CD algorithm and that
in SB-GAM is that SB-GAM uses a blocked CD algorithm for their group
prior, while the proposed prior is pairwise independent requiring no
special treatment in the CD algorithm.

@@

Similar to the EM-CD algorithm, the EM-IWLS algorithm is an iterative
EM-based algorithm where the iterative weighted least squares algorithm
is used to find the estimate of \(\boldsymbol{\beta}, \phi\) that
maximizes \(E(Q_1)\). The iterative weighted least squares algorithm was
originally proposed to fit the classical generalized linear models, and
generalized to fit some Bayesian hierarchical models.\citep{Gelman2013}
Yi and Ma \citep{Yi2012} formulated Student's t-distribution and double
exponential distribution as hierarchical normal distributions such that
generalized linear models with shrinkage priors can be easily fitted
using IWLS in combination with EM algorithm. In this work, we adapt the
EM-IWLS paradigm to fit BHAM with spike-and-slab spline prior, with the
full potential to adopt Student's t, double exponential and mixture
Student's t priors when sparse assumption is not necessary.

>>>>

\address[a]{Department of Biostatistics, University of Alabama at
Birmingham, Birmingham, USA}
\address[b]{Department of Biostatistics and Data Science, Wake Forest
School of Medicine, Winston-Salem, USA}

@@

\abstract{There are proposals that extend the classical generalized
additive models (GAMs) to accommodate high-dimensional data (\(p>>n\))
using group sparse regularization. However, the sparse reuglarization
may induce excess shrinkage when estimating smoothing functions,
damaging predictive performance. Moreover, most of these GAMs consider
an ``all-in-all-out'' approach for function selection, rendering them
difficult to answer if nonlinear effects are necessary. While some
Bayesian models can address these shortcomings, using Markov chain Monte
Carlo (MCMC) algorithms for model fitting creates a new challenge,
scalability. Hence, we propose Bayesian hierarchical generalized
additive models as a solution: we incorporate the smoothing penalty that
can separate linear and nonlinear spaces of smoothing function. A novel
spike-and-slab spline prior is used to select the components of
smoothing functions. Two scalable and deterministic algorithms,
EM-Coordinate Descent and EM-Iterative Weighted Least Squares, are
developed for different utilities. Simulation studies and metabolomics
data analyses demonstrate improved predictive or computational
performance against state-of-the-art models, mgcv, COSSO and sparse
Bayesian GAM. The software implementation of the proposed models is
freely available via an R package BHAM.}

@@

Many modern biomedical research, e.g.~sequencing data analysis, electric
health record data analysis, require special treatment of
high-dimensionality, commonly known as \(p >> n\) problem. There is
extensive literature on high-dimensional linear models via penalized
models or Bayesian hierarchical models, see Mallick and Yi
\citep{Mallick2013} for review. These models are built upon a
restrictive and unrealistic assumption, linearity. In classical
statistical modeling, many strategies and models are proposed to relax
the linearity assumption with various degrees of complexity. For
example, variable categorization is a simple and common practice in
epidemiology, but suffers from power and interpretation issues. More
complex models to address nonlinear effects include random forest and
other so-called ``black box'' models \citep{Breiman2001}. These models
are useful for statistical prediction but do not estimate parameters
relevant to the data generation process that one can draw inferences
from. In addition, how to generalize these models to the
high-dimensional setting remains unclear.

Nonparametric regression models are appropriate alternatives to the
``black-box'' models thanks to their balance between model flexibility
and interpretability. Among those, generalized additive models (GAMs),
proposed in the seminal work of Hastie and Tibshirani
\citep{Hastie1987}, grew to be one of the most popular modeling tools.
In a GAM, the response variable, \(Y\), which is assumed to follow some
exponential family distribution with mean \(\mu\) and dispersion
\(\phi\), can be modeled with the summation of smoothing functions,
\(B_j(\cdot), j = 1, \dots, p\), of a given \(p\)-dimensional vector of
covariates \(\boldsymbol{x}\), written as \[
 E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)),
\] where \(g^{-1}(\cdot)\) is the inverse of a monotonic link function.
The smoothing functions can take many forms and are estimated using a
pseudo-weighted version of the backfitting algorithm
\citep{Breiman1985}. Nevertheless, the classical GAMs cannot fulfill the
increasing analytic demands for high-dimensional data analysis in
biomedical studies.

There exists some proposals to generalize the classical GAM to
accommodate high-dimensional applications. The regularized models,
branching out from group regularized linear models, are used to fit GAMs
by accounting for the structure introduced when expanding smoothing
functions. Ravikumar et al. \citep{Ravikumar2009} extended the grouped
lasso \citep{Yuan2006} to additive models (AMs); Huang et al.
\citep{Huang2010} further developed adaptive grouped lasso for additive
models; Wang et al. \citep{Wang2007} and Xue \citep{Xue2009}
respectively applied grouped SCAD penalty \citep{Fan2001} to additive
models. Recent Bayesian hierarchical models are also used in the context
of high-dimensional additive models. Various group spike-and-slab priors
combining with computationally intensive Markov chain Monte Carlo (MCMC)
algorithms \citep{Xu2015, Yang2020} are proposed, where the application
on AMs are by-products. Bai et al. \citep{Bai2020} was the first to
apply group spike-and-slab lasso prior to Gaussian AMs using a fast
optimization algorithm, and further generalized the framework to
GAMs\citep{Bai2021}. Focus on addressing the sparsity, these methods can
overly penalize the basis function coefficients and produce inaccurate
predictions and curve interpolation, particularly when complex signals
are assumed and large number of knots are used. \citep{Scheipl2013} In
addition, these methods adapt an `all-in-all-out' strategy, i.e.~either
including or excluding the variable completely, rendering no space for
bi-level selection. Scheipl et al. \citep{Scheipl2012} proposed a
spike-and-slab structure prior, normal-mixture-of-inverse gamma prior,
that address the previous challenges. But the model fitting relies on
computational intensive MCMC algorithmic and creates scalability
concern. It would be of special interest to develop a fast, flexible and
accurate generalized additive model framework.

@@

The proposed framework, BHAM, differs from previous spike-and-slab based
GAMs, i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB-GAM
\citep{Bai2021} in three ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM.
Spike-and-slab lasso priors provide computational convenience during
model fitting by using optimization algorithms instead of computational
intensive sampling algorithms. They make fitting high- and
ultrahigh-dimensional models more feasible without sacrificing
performance in prediction and variable selection. Secondly, the
optimization algorithms introduced in this article include EM-Coordinate
Descent (EM-CD) algorithm and EM-Iterative weighted least square
(EM-IWLS) algorithm for different utilities. The applications of the two
aforementioned algorithms are fruitful in Bayesian hierarchical models
for high-dimensional data. \citep{Yi2012, Rockova2014a, Rockova2018}
Even though SB-GAM with a group spike-and-slab lasso prior also uses
EM-CD for model fitting, the proposed EM-IWLS provides
variance-covariance matrix of the coefficients, which has the potential
to provide uncertainty inference. EM-IWLS complements the lack of
inference potential from EM-CD in a median and high dimensional setting.
Furthermore, the proposed model addresses the incapability of bi-level
selection in SB-GAM. Lastly, the R implementation of the Bayesian
hierarchical additive model is available publicly via BHAM at
\url{https://github.com/boyiguo1/BHAM}, making translational science
more accessible.

In Section \ref{sec:BHAM}, we establish the Bayesian hierarchical
generalized additive model, introduce the proposed spike-and-slab spline
priors, and describe the two fast-fitting algorithms. In Section
\ref{sec:sim}, we compare the proposed framework to state-of-the-art
models, mgcv, COSSO and sparse Bayesian GAM via Monte Carlo simulation
studies. Analyses of two metabolomics datasets are presented in Section
\ref{sec:real_data}. Conclusion and discussions are given in Section
\ref{sec:concl}.

@@

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model with link function \(g(\cdot)\) and
linear predictor \begin{equation}\label{eq:gam}
\eta = \beta_0 + \sum\limits^p_{j=1}B_j(x_j) = \beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j,
\end{equation} with smoothing functions \(B_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The outcome \(Y\) follows a exponential family
distribution with density function \(f(y)\), mean \(\mu = g^{-1}(\eta)\)
and dispersion parameter \(\phi\), and the data distribution is
\begin{equation}
f(\boldsymbol{Y} = \boldsymbol{y}| \boldsymbol{\beta}, \phi) = \prod\limits^n_{i=1}f( Y = y_i|\boldsymbol{\beta}, \phi),\nonumber
\end{equation} The basis function matrix, i.e.~the design matrix derived
from the smoothing function \(B_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+3\)
elements column vector. Its corresponding smoothing penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable such that
\(B_j(x_j) = \boldsymbol{\beta}_j^T \boldsymbol{X}_j\). With slight
abuse of notation, we denote vectors and matrices in bold fonts
\(\boldsymbol{\beta}, \boldsymbol{X}\) with conformable dimensions,
where scalar and random variables are denoted in with unbold fonts
\(\beta, X\). The matrix transposing operation is denoted with a
superscript \(^T\). To note, the proposed model can include parametric
forms of variables in the model, treating the parametric function as a
special case of the smoothing function, e.g.~\(B_j(x_j) = x_j\) with the
smoothing penalty matrix defined as
\(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

@@

The re-parameterization step sets up the foundation of the proposed
model and provides three benefits. First of all, the re-parameterization
integrates the smoothing penalty into the design matrix, and encourages
models to properly smooth the nonlinear function in addition to the
sparse penalty for function selection. Secondly, the eigendecomposition
of the smoothing penalty allows the isolation of the linear from the
nonlinear space, improving the feasibility of bi-level function
selection and inference. The eigendecomposition facilitates the
construction of orthonormal design matrix, which makes imposing
independent priors on the coefficients possible. This reduces the
computational complexity compared to using a multivariate priors, and
greatly broadens the choices of priors and further model choices. Last
but not least, the scaling step of \(\boldsymbol{X}^*\) further
simplifies the choice of priors, from column-specific priors to a
unified prior.

\hypertarget{spike-and-slab-spline-priors}{%
\subsection{Spike-and-Slab Spline
Priors}\label{spike-and-slab-spline-priors}}

The family of spike-and-slab (SS) priors regression models is one of
most commonly used models in high-dimensional data analysis for its
utility in outcome prediction and variable selection. We defer to Bai et
al. \citep{Bai2021Review} for an in-depth introduction to spike-and-slab
priors. To summarize, spike-and-slab priors are a family of mixture
distributions that comprises a skinny spike density
\(f_{\text{spike}}(\cdot)\) for weak signals and a flat slab density
\(f_{\text{slab}}(\cdot)\) for strong signals, mathematically \[
 \beta|\gamma \sim (1-\gamma)f_{\text{spike}}(\beta) + \gamma f_{\text{slab}}(\beta).
\] The most distinct feature of SS priors is that it is conditioned on a
latent binary variable \(\gamma \in \{0,1\}\) that indicates whether the
variable \(x\) is included in the model. There are various
spike-and-slab priors depending on the choice for the spike density
\(f_{\text{spike}}(\cdot)\) and the slab density
\(f_{\text{slab}}(\cdot)\), see George and McCulloch
\citep{George1993, George1997}; Chipman \citep{Chipman1996} for grouped
variables; Brown et al. \citep{Brown1998} for multivariate outcomes;
Ishwaran and Rao \citep{Ishwaran2005}; Clyde and George
\citep{Clyde2004} and reference therein. Two of most popular spike and
slab priors are spike-and-slab normal prior \citep{George1993} and
spike-and-slab double exponential prior\citep{Rockova2018}.

The major criticism of early spike-and-slab models is being
computationally prohibitive. \citep{Bai2021Review} Since then, many
studies focus on alleviating the computational burden that sampling
algorithms bear, which include EMVS based on spike-and-slab normal
prior\citep{Rockova2014a} and spike-and-slab Lasso
\citep{Rockova2018b, Rockova2018}. Particularly, the development of
spike-and-slab Lasso substantially improves the scalability of SS
models, setting up the theoretical foundation for generalized models in
-omics data analysis \citep{Tang2017a, Tang2017, Tang2018, Tang2019}.
The spike-and-slab Lasso (SSL) prior is composed of two double
exponential distributions with mean 0 and different dispersion
parameters, \(0 < s_0 < s_1\), mathematically, \begin{equation} 
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.\nonumber
\end{equation} Given that both double exponential distributions have a
mean of 0 and the latent indicator \(\gamma\) can only take the value of
0 or 1, the mixture double exponential distribution can be formulated as
one single double exponential density, \begin{equation} \label{eq:ssl}
\beta | \gamma \sim DE(0, S), 0 < s_0 < s_1,
\end{equation} with the scale parameter
\(S = (1-\gamma)s_0 + \gamma s_1\). The SSL also mitigates the problem
of EMVS where the weak signals are not shrink to zero, and hence is
preferred in high-dimensional data analysis. We notice that Bai
\citep{Bai2021} is the first to apply spike-and-slab lasso prior in the
GAM framework, where the densities of the spike and slab components take
the group lasso density \citep{Xu2015}. Bai \citep{Bai2021} takes an
``all-in-all-out'' strategy for function selection.

\hypertarget{two-part-spike-and-slab-lasso-prior}{%
\subsubsection{Two-part Spike-and-Slab Lasso
Prior}\label{two-part-spike-and-slab-lasso-prior}}

We introduce a novel prior for GAMs, particularly for high-dimensional
nonlinear modeling with bi-level selection. The proposed prior extends
from the spike-and-slab lasso prior described in Equation
(\ref{eq:ssl}). Given the re-parameterized design matrix
\(\boldsymbol{X}_j = \begin{bmatrix} X^0_j : \boldsymbol{X}_j^*\end{bmatrix}\)
for the \(j\)th variable, we impose a two-part SSL prior to the
coefficients
\(\boldsymbol{\beta}_j = \begin{bmatrix} \beta_j : \boldsymbol{\beta}_j^*\end{bmatrix}\).
Specifically, we impose independent group priors on the linear space
coefficients and on the nonlinear space coefficients respectively,
\begin{align}\label{eq:bham_ssl}
  \beta_{j} |\gamma_{j},s_0,s_1 &\sim DE(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1) \nonumber \\
  \beta^*_{jk} | \gamma^*_{j},s_0,s_1 &\overset{\text{iid}}{\sim}DE(0,(1-\gamma^*_{j}) s_0 + \gamma^*_{j} s_1), k=1,\dots, K_j
\end{align} where \(\gamma_{j}\in\{0,1\}\) and
\(\gamma^*_{j}\in \{0,1\}\) are two latent indicator variables,
indicating if the model includes the linear effect and the nonlinear
effect of the \(j\)th variable respectively. \(s_0\) and \(s_1\) are
scale parameters, assuming \(0 < s_0 < s_1\) and given. These scale
parameters \(s_0\) and \(s_1\) can be treated as tuning parameters and
optimized via cross-validation. A discussion of how to choose the scale
parameters comes in Section \ref{sec:tune}. To note, this prior differs
from previous group spike-and-slab lasso priors
\citep{Tang2018, Tang2019}, as the \(\beta_j\) and
\(\boldsymbol{\beta}^*_j\) have different indicator variables
\(\gamma_j\), \(\gamma_j^*\) respectively. It is possible to add a more
restrictive assumption on the priors, assuming that one indicator
variable decides the inclusion of both the linear effect and nonlinear
effect, i.e.~\(\gamma_j = \gamma^*_j\). This converges to the SB-GAM
\citep{Bai2021}. Conversely, it is also possible to relax the assumption
such that each coefficient \(\beta_{jk} \in \boldsymbol{\beta}^*_j\) has
its own latent indicator \(\gamma_{jk}\), but at the cost of
complicating the bi-level function selection. This reduces the proposed
prior to the classic spike-and-slab lasso prior.

The re-parameterization introduced in Section \ref{sec:BHAM} grants the
validity of the proposed prior. First of all, the smoothing function
bases are linear dependent and necessitate extra attention. The
eigeondecomposition remedies the problem and hence our prior can be set
to be conditionally independent. Secondly, the eigenvalue scaling
provides a panacea to allow unified scale parameters for all bases of
all smoothing functions of variables.

@@

\begin{center}
Figure \ref{fig:SSprior} here
\end{center}

@@

In this section, we extend the two EM-based algorithms, EM-CD and
EM-IWLS algorithms, to fit BHAMs. To note, the two proposed algorithms
provides different utilities. The EM-CD algorithm is specifically for
fitting BHAM-SSL with an expedited performance, recommending to use in
high and ultra-high dimensional setting. A specific concern of EM-CD
algorithm is that it provides no information for inference. In contrast,
the EM-IWLS can estimate the variance-covariance matrix of the
coefficients. Moreover, the EM-IWLS is a more general model fitting
algorithm that can be used for fitting not only SSL and continuous SS
priors but also Student's t-priors and double exponential priors. To
note, SB-GAM\citep{Bai2020, Bai2021} also used an EM-CD algorithm. The
main difference between proposed EM-CD algorithm and that in SB-GAM is
that SB-GAM uses a blocked CD algorithm for their group prior, while the
proposed prior is pairwise independent requiring no special treatment in
the CD algorithm.

@@

Similar to the EM-CD algorithm, the EM-IWLS algorithm is an iterative
EM-based algorithm where the iterative weighted least squares algorithm
is used to find the estimate of \(\boldsymbol{\beta}, \phi\) that
maximizes \(E(Q_1)\). The iterative weighted least squares algorithm was
originally proposed to fit the classical generalized linear models, and
generalized to fit some Bayesian hierarchical models.\citep{Gelman2013}
Yi and Ma \citep{Yi2012} formulated Student's t-distribution and double
exponential distribution as hierarchical normal distributions such that
generalized linear models with shrinkage priors can be easily fitted
using IWLS in combination with EM algorithm. In this work, we adapt the
EM-IWLS paradigm to fit BHAM with spike-and-slab spline prior .
